---
title: "final_eda"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(root.dir ="/Users/arianaschindler/Documents/2023_spring/DATA2020/final_project")
```

```{r}
# clearing environment
rm(list=ls())

# loading in the cleaned data
wd <- "~/Desktop/DSI Spring/Stats/DATA-2020/"

setwd(wd)
load("cleaned_data.Rdata")


# loading packages
library(ggplot2)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(gridExtra)

# turning race feature into categorical factor
df <- df %>%
  mutate(race = ifelse(race %in% c(".n:  No answer" ,
                                 ".s:  Skipped on Web", 
                                 ".i:  Inapplicable",
                                 ".d:  Do not Know/Cannot Choose"), "missing", race)) %>%
  mutate(race = as.factor(race))

# turning zodiac feature into categorical factor
df <- df %>%
  mutate(zodiac = ifelse(zodiac %in% c(".n:  No answer" ,
                                   ".s:  Skipped on Web", 
                                   ".i:  Inapplicable",
                                   ".d:  Do not Know/Cannot Choose"), "missing", zodiac)) %>%
  mutate(zodiac = as.factor(zodiac))

# turning nulls into "No Response" for possible target features
df <- df %>%
  mutate(wealth_imp = ifelse(is.na(wealth_imp), "No Response", as.character(wealth_imp))) %>%
  mutate(wealth_imp = factor(wealth_imp))

df <- df %>%
  mutate(parents_imp = ifelse(is.na(parents_imp), "No Response", as.character(parents_imp))) %>%
  mutate(parents_imp = factor(parents_imp))

df <- df %>%
  mutate(educ_imp = ifelse(is.na(educ_imp), "No Response", as.character(educ_imp))) %>%
  mutate(educ_imp = factor(educ_imp))

df <- df %>%
  mutate(hardWork_imp = ifelse(is.na(hardWork_imp), "No Response", as.character(hardWork_imp))) %>%
  mutate(hardWork_imp = factor(hardWork_imp))

df <- df %>%
  mutate(rightPpl_imp = ifelse(is.na(rightPpl_imp), "No Response", as.character(rightPpl_imp))) %>%
  mutate(rightPpl_imp = factor(rightPpl_imp))

df <- df %>%
  mutate(political_imp = ifelse(is.na(political_imp), "No Response", as.character(political_imp))) %>%
  mutate(political_imp = factor(political_imp))

df <- df %>%
  mutate(race_imp = ifelse(is.na(race_imp), "No Response", as.character(race_imp))) %>%
  mutate(race_imp = factor(race_imp))

df <- df %>%
  mutate(religion_imp = ifelse(is.na(religion_imp), "No Response", as.character(religion_imp))) %>%
  mutate(religion_imp = factor(religion_imp))

df <- df %>%
  mutate(sex_imp = ifelse(is.na(sex_imp), "No Response", as.character(sex_imp))) %>%
  mutate(sex_imp = factor(sex_imp))


# splitting data into two dfs for each year
df1 <- df %>%
  filter(df$year=='1987')
df2 <- df %>%
  filter(df$year=='2021')
```

```{r}
# extracting dimensions for rows and columns
# 1987 data
dim_rows1 <- dim(df1)[1]
dim_cols1 <- dim(df1)[2]

#2021 data
dim_rows2 <- dim(df2)[1]
dim_cols2 <- dim(df2)[2]

# printing dimensions
cat(paste('Number of Rows in Survey Data from 1987:', dim_rows1, '\nNumber of Columns:', dim_cols1, '\n', '\nNumber of Rows in Survey Data from 2021:', dim_rows2, '\nNumber of Columns:', dim_cols2))
```


```{r}
#==========================================
# CHECKING OUT POSSIBLE TARGET FEATURES
#==========================================

# function to calculate the proportions of any categorical feature


compute_props <- function(df, group_col, weight_col) {
  d <- {{df}} %>%
    group_by({{ group_col }}, year) %>%
    summarize(prop = sum({{ weight_col }}, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(prop = prop / sum(prop))
  
  return(d)
}

# function to make year comparison 

make_year_graph <- function(m) {
  g <- ggplot(prop_df, aes(fill={{m}}, y=prop, x=year)) + 
        geom_bar(position="stack", stat="identity") +
        geom_text(aes(label = scales::percent(prop)), 
                  position = position_stack(vjust = 0.5), 
                  size = 2) +
        labs(title = deparse(substitute(m)),
             x = "Response",
             y = "Proportion") + 
        theme(text = element_text(size = 8))
}

# wealth_imp ------------

# calculate weighted proportion by group

prop_df1 <- compute_props(df1, wealth_imp, wgt_comb) 
prop_df2 <- compute_props(df2, wealth_imp, wgt_comb)

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_wealth_imp <- make_year_graph(wealth_imp)
plot_wealth_imp

ggsave("EDA Images/wealth_time.png", device = "png")


# parents_imp ------------
prop_df1 <- compute_props(df1, parents_imp, wgt_comb) 
prop_df2 <- compute_props(df2, parents_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_parents_imp <- make_year_graph(parents_imp)
plot_parents_imp
ggsave("EDA Images/parents_time.png", device = "png")

# educ_imp ------------
prop_df1 <- compute_props(df1, educ_imp, wgt_comb) 
prop_df2 <- compute_props(df2, educ_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_educ_imp <- make_year_graph(educ_imp)
plot_educ_imp
ggsave("EDA Images/educ_time.png", device = "png")


# hardWork_imp ------------
prop_df1 <- compute_props(df1, hardWork_imp, wgt_comb) 
prop_df2 <- compute_props(df2, hardWork_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_hardWork_imp <- make_year_graph(hardWork_imp)
plot_hardWork_imp
ggsave("EDA Images/hardwork_time.png", device = "png")


# rightPpl_imp ------------
prop_df1 <- compute_props(df1, rightPpl_imp, wgt_comb) 
prop_df2 <- compute_props(df2, rightPpl_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_rightPpl_imp <- make_year_graph(rightPpl_imp)
plot_rightPpl_imp
ggsave("EDA Images/rightPpl_imp_time.png", device = "png")



# political_imp ------------
prop_df1 <- compute_props(df1, political_imp, wgt_comb) 
prop_df2 <- compute_props(df2, political_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_political_imp <- make_year_graph(political_imp)
plot_political_imp
ggsave("EDA Images/political_imp_time.png", device = "png")


# race_imp ------------
prop_df1 <- compute_props(df1, race_imp, wgt_comb) 
prop_df2 <- compute_props(df2, race_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_race_imp <- make_year_graph(race_imp)
plot_race_imp
ggsave("EDA Images/race_imp_time.png", device = "png")


# religion_imp ------------
prop_df1 <- compute_props(df1, religion_imp, wgt_comb) 
prop_df2 <- compute_props(df2, religion_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_religion_imp <- make_year_graph(religion_imp)
plot_religion_imp
ggsave("EDA Images/religion_imp_time.png", device = "png")


# sex_imp ------------
prop_df1 <- compute_props(df1, sex_imp, wgt_comb) 
prop_df2 <- compute_props(df2, sex_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 , prop_df2)

# stacked + percent barchart
plot_sex_imp <- make_year_graph(sex_imp)
plot_sex_imp
ggsave("EDA Images/sex_imp_time.png", device = "png")


# arrange the plots in a 3 by 3 grid using arrangeGrob
targets_w_na <- grid.arrange(arrangeGrob(plot_wealth_imp, plot_parents_imp, plot_educ_imp, ncol = 3),
             arrangeGrob(plot_hardWork_imp, plot_rightPpl_imp, plot_political_imp, ncol = 3),
             arrangeGrob(plot_race_imp, plot_religion_imp, plot_sex_imp, ncol = 3),
             nrow = 3)

ggsave("EDA Images/Combined Year Plot.png", plot = targets_w_na)
```

We can see from the grid above that the percentage of people who took the survey in 1987 did not respond to a given question was about 14-17%, and that percentage shot up to about 53-55% in 2021. Next, we can account for this to see if any features had a drastic difference between responses from 1987 and 2021.

```{r}
# wealth_imp ------------
sub_df1 <- subset(df1, wealth_imp != "No Response")
sub_df2 <- subset(df2, wealth_imp != "No Response")

prop_df1 <- compute_props(sub_df1, wealth_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, wealth_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_wealth_imp <- make_year_graph(wealth_imp)
plot_wealth_imp

ggsave("EDA Images/wealth_time_nomissing.png", device = "png")

# parents_imp ------------
sub_df1 <- subset(df1, parents_imp != "No Response")
sub_df2 <- subset(df2, parents_imp != "No Response")

prop_df1 <- compute_props(sub_df1, parents_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, parents_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_parents_imp <- make_year_graph(parents_imp)
plot_parents_imp

ggsave("EDA Images/parents_imp_nomissing.png", device = "png")


# educ_imp ------------
sub_df1 <- subset(df1, educ_imp != "No Response")
sub_df2 <- subset(df2, educ_imp != "No Response")

prop_df1 <- compute_props(sub_df1, educ_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, educ_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_educ_imp <- make_year_graph(educ_imp)
plot_educ_imp

ggsave("EDA Images/educ_imp_nomissing.png", device = "png")


# hardWork_imp ------------
sub_df1 <- subset(df1, hardWork_imp != "No Response")
sub_df2 <- subset(df2, hardWork_imp != "No Response")

prop_df1 <- compute_props(sub_df1, hardWork_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, hardWork_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_hardWork_imp <- make_year_graph(hardWork_imp)
plot_hardWork_imp

ggsave("EDA Images/hardWork_imp_nomissing.png", device = "png")



# rightPpl_imp ------------
sub_df1 <- subset(df1, rightPpl_imp != "No Response")
sub_df2 <- subset(df2, rightPpl_imp != "No Response")

prop_df1 <- compute_props(sub_df1, rightPpl_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, rightPpl_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_rightPpl_imp <- make_year_graph(rightPpl_imp)
plot_rightPpl_imp

ggsave("EDA Images/hardWork_imp_nomissing.png", device = "png")


# political_imp ------------
sub_df1 <- subset(df1, political_imp != "No Response")
sub_df2 <- subset(df2, political_imp != "No Response")

prop_df1 <- compute_props(sub_df1, political_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, political_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_political_imp <- make_year_graph(political_imp)
plot_political_imp

ggsave("EDA Images/political_imp_nomissing.png", device = "png")



# race_imp ------------
sub_df1 <- subset(df1, race_imp != "No Response")
sub_df2 <- subset(df2, race_imp != "No Response")

prop_df1 <- compute_props(sub_df1, race_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, race_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_race_imp <- make_year_graph(race_imp)
plot_race_imp

ggsave("EDA Images/race_imp_nomissing.png", device = "png")



# religion_imp ------------
sub_df1 <- subset(df1, religion_imp != "No Response")
sub_df2 <- subset(df2, religion_imp != "No Response")

prop_df1 <- compute_props(sub_df1, religion_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, religion_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_religion_imp <- make_year_graph(religion_imp)
plot_religion_imp

ggsave("EDA Images/religion_imp_nomissing.png", device = "png")



# sex_imp ------------
sub_df1 <- subset(df1, sex_imp != "No Response")
sub_df2 <- subset(df2, sex_imp != "No Response")

prop_df1 <- compute_props(sub_df1, sex_imp, wgt_comb) 
prop_df2 <- compute_props(sub_df2, sex_imp, wgt_comb) 

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

prop_df <- bind_rows(prop_df1 , prop_df2)

plot_sex_imp <- make_year_graph(sex_imp)
plot_sex_imp

ggsave("EDA Images/sex_imp_nomissing.png", device = "png")



# arrange the plots in a 3 by 3 grid using arrangeGrob
targets_wo_na <- grid.arrange(arrangeGrob(plot_wealth_imp, plot_parents_imp, plot_educ_imp, ncol = 3),
             arrangeGrob(plot_hardWork_imp, plot_rightPpl_imp, plot_political_imp, ncol = 3),
             arrangeGrob(plot_race_imp, plot_religion_imp, plot_sex_imp, ncol = 3),
             nrow = 3)
ggsave("EDA Images/targets_without_missing.png", plot = targets_wo_na)
```

This gives us a much better view of what general public opinions were of different importance features during the two recorded years. We can see in features such as "Parent Education Importance", "Right People Importance", and "Hard Work Importance" that the opinions between the two time periods have not changed much. We start to see a little more difference in responses when we look at features like "Wealth Importance", where in 1987, there was only a ~3% difference in the proportion of people agreeing and disagreeing with the importance feature, whereas in 2021, the difference in proportions is just over 20%, with the majority of responses answering that they believed that having a wealthy family is necessary to get ahead in life. We can see similar differences in the features "Religion Importance", "Political Importance", and "Race Importance".

Now we can check out the distributions of other features in our data.

```{r, out.width="50%", out.height="50%", include=TRUE, fig.align="center"}
# age distribution for different years
df1_no_na <- drop_na(df1)
df2_no_na <- drop_na(df2)

df1_age <- ggplot(data = df1_no_na, aes(x = age)) + 
  geom_histogram(bins = 30) +
  labs(title = "Histogram of Respondent Age in 1987", x = "Respondent Age", y = "Frequency") +
  theme(text = element_text(size = 8))

df2_age <- ggplot(data = df2_no_na, aes(x = age)) + 
  geom_histogram(bins = 30) +
  labs(title = "Histogram of Respondent Age in 2021", x = "Respondent Age", y = "Frequency") +
  theme(text = element_text(size = 8))

# arrange the plots in a 3 by 3 grid using arrangeGrob
age_dist_hist <- grid.arrange(arrangeGrob(df1_age, df2_age, ncol = 2))

#ggsave("age_histogram.png", plot = age_dist_hist, width=10, height=6)
```


```{r}
summary(df1$age)
```

```{r}
summary(df2$age)
```

Through analysis, we need to keep in mind that while there are 34 years between the survey dates, there is a 12 year difference in the median age of respondents and a 7 year difference in the mean age of respondents, despite the max and min being the same in both years.

```{r}
df <- drop_na(df)
age_boxplot <- ggplot(df, aes(x=factor(year), y=age)) + 
  geom_boxplot() + 
  labs(x = "Year", y = "Age", title='Boxplot Distributions of Age in Varying Years') +
  theme(text = element_text(size = 8))

age_boxplot
#ggsave("age_boxplot.png", plot = age_boxplot, width=10, height=6)
```

```{r}
# education distribution for different years
df1_educ <- ggplot(data = df1_no_na, aes(x = educ)) + 
  geom_histogram(bins = 30) +
  labs(title = "Histogram of Respondent Education in 1987", x = "Years of Respondent Education", y = "Frequency") +
  theme(text = element_text(size = 8))

df2_educ <- ggplot(data = df2_no_na, aes(x = educ)) + 
  geom_histogram(bins = 30) +
  labs(title = "Histogram of Respondent Education in 2021", x = "Years of Respondent Education", y = "Frequency") +
  theme(text = element_text(size = 8))

# arrange the plots in a 2 by 2 grid using arrangeGrob
 educ_hist <- grid.arrange(arrangeGrob(df1_educ, df2_educ, ncol = 2))
 
#ggsave("educ_histograms.png", plot = educ_hist, width=10, height=6)
```

```{r}
# work status distribution
df1_wrkstat <- ggplot(data = df1_no_na, aes(x = wrkstat)) +
  geom_bar(fill = "tomato4") +
  labs(title="Respondent Work Status in 1987", 
       x="Work Status", y="Count") +
  scale_x_discrete(labels = c("In school" = "In School",
                             "Keeping house" = "Keeping house",
                             "missing" = "Missing",
                             "Other" = "Other",
                             "Retired" = "Retired",
                             "Unemployed, laid off, looking for work" = "Unemployed",
                             "With a job, but not at work because of temporary illness, vacation, strike" = "Not Working",
                             "Working full time" = "Full Time",
                             "Working part time" = "Part Time")) +
  coord_flip() +
  theme(text = element_text(size = 8))

df2_wrkstat <- ggplot(data = df2_no_na, aes(x = wrkstat)) +
  geom_bar(fill = "tomato4") +
  labs(title="Respondent Work Status in 2021", 
       x="Work Status", y="Count") +
  scale_x_discrete(labels = c("In school" = "In School",
                             "Keeping house" = "Keeping house",
                             "missing" = "Missing",
                             "Other" = "Other",
                             "Retired" = "Retired",
                             "Unemployed, laid off, looking for work" = "Unemployed",
                             "With a job, but not at work because of temporary illness, vacation, strike" = "Not Working",
                             "Working full time" = "Full Time",
                             "Working part time" = "Part Time")) +
  coord_flip() +
  theme(text = element_text(size = 8))

# arrange the plots in a 2 by 2 grid using arrangeGrob
wrkstat_hist <- grid.arrange(arrangeGrob(df1_wrkstat, ncol = 1),
             arrangeGrob(df2_wrkstat, ncol = 1))

#ggsave("wrkstat_histogram.png", plot = wrkstat_hist)
```

```{r}
prop_df1 <- compute_props(df1_no_na, wrkstat, wgt_comb)
prop_df2 <- compute_props(df2_no_na, wrkstat)

prop_df <- bind_rows(prop_df1 , prop_df2 )

# stacked + percent barchart
wrkstat_fig <- ggplot(prop_df, aes(fill=variable, y=prop, x=wrkstat)) + 
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = scales::percent(prop)), 
            position = position_stack(vjust = 0.5), 
            size = 2) +
  labs(title = "Work Status of Respondents in 1987 vs 2021",
       x = "Status",
       y = "Proportion") + 
  scale_x_discrete(labels = c("In school" = "In School",
                             "Keeping house" = "Keeping house",
                             "missing" = "Missing",
                             "Other" = "Other",
                             "Retired" = "Retired",
                             "Unemployed, laid off, looking for work" = "Unemployed",
                             "With a job, but not at work because of temporary illness, vacation, strike" = "Not Working",
                             "Working full time" = "Full Time",
                             "Working part time" = "Part Time")) +
  theme(text = element_text(size = 8))

wrkstat_fig
#ggsave("wrkstat_stacked_props.png", plot = wrkstat_fig, width=10, height=6)
#levels(df$wrkstat)
```

One interesting observation at this point is if we look at the distribution of age, the survey in 2021 had a much higher number of retirement-aged respondents which helps explain the large difference in proportion of retired respondents between the two timeframes.

```{r}
prop_df1 <- compute_props(df1_no_na, marital)
prop_df2 <- compute_props(df2_no_na, marital)

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

# stacked + percent barchart
marital_fig <- ggplot(prop_df, aes(fill=variable, y=prop, x=marital)) + 
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = scales::percent(prop)), 
            position = position_stack(vjust = 0.5), 
            size = 2) +
  labs(title = "Marital Status of Respondents in 1987 vs 2021",
       x = "Status",
       y = "Proportion") + 
  theme(text = element_text(size = 8))

marital_fig
#ggsave("marital_dist.png", plot = marital_fig)
#levels(df$marital)
```

```{r}
prop_df1 <- compute_props(df1_no_na, sex)
prop_df2 <- compute_props(df2_no_na, sex)

prop_df <- bind_rows(prop_df1 %>% mutate(variable = "1987"),
                         prop_df2 %>% mutate(variable = "2021"))

# stacked + percent barchart
sex_proportions <- ggplot(prop_df, aes(fill=variable, y=prop, x=sex)) + 
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = scales::percent(prop)), 
            position = position_stack(vjust = 0.5), 
            size = 2) +
  labs(title = "Sex Proportion of Respondents in 1987 vs 2021",
       x = "Response",
       y = "Proportion") + 
  theme(text = element_text(size = 8))

sex_proportions
ggsave("sex_proportions.png", plot = sex_proportions, width=8, height=6)
#levels(df$marital)
```